---
output: 
  github_document
---

# ospsuite.parameteridentification


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  message = FALSE,
  warning = FALSE,
  out.width = "100%"
)
```


  <!-- badges: start -->

[![Main-Workflow](https://img.shields.io/github/actions/workflow/status/Open-Systems-Pharmacology/OSPSuite.ParameterIdentification/Main-Workflow.yaml?branch=main&label=Build)](https://github.com/Open-Systems-Pharmacology/OSPSuite.ParameterIdentification/actions/workflows/Main-Workflow.yaml)
[![codecov](https://codecov.io/gh/Open-Systems-Pharmacology/OSPSuite.ParameterIdentification/graph/badge.svg?token=FL1AEQ7316)](https://codecov.io/gh/Open-Systems-Pharmacology/OSPSuite.ParameterIdentification)

  <!-- badges: end -->
  
<!-- README.md is generated from README.Rmd. Please edit that file -->


The **ospsuite.parameteridentification** R package provides the functionality of performing parameter identification (i.e., fitting the model to observed data) with simulations created in the Open Systems Pharmacology Software tools PK-Sim and MoBi. The package requires the [**ospsuite**](https://github.com/Open-Systems-Pharmacology/OSPSuite-R) package to run the simulations.

## Installation

The package can be installed from GitHub using the `{pak}` package. Simply run the following code:

```{r, eval=FALSE}
install.packages("pak")

pak::pak("Open-Systems-Pharmacology/OSPSuite.ParameterIdentification@*release")
```

Get the latest development version with:

```{r, eval=FALSE}
pak::pak("Open-Systems-Pharmacology/OSPSuite.ParameterIdentification")
```


## User guide

Examples of running parameter estimation tasks using this package are detailed in `vignette('user-guide')`.


## Code of conduct

Everyone interacting in the Open Systems Pharmacology community (codebases, issue trackers, chat rooms, mailing lists etc...) is expected to follow the Open Systems Pharmacology [code of conduct](https://github.com/Open-Systems-Pharmacology/Suite/blob/master/CODE_OF_CONDUCT.md).

## Contribution

We encourage contribution to the Open Systems Pharmacology community. Before getting started please read the [contribution guidelines](https://github.com/Open-Systems-Pharmacology/Suite/blob/master/CONTRIBUTING.md). If you are contributing code, please be familiar with the [coding standards](https://github.com/Open-Systems-Pharmacology/Suite/blob/master/CODING_STANDARDS_R.md).

## License

The `{OSPSuite.ParameterIdentification}` package is released under the [GPLv2 License](LICENSE).

All trademarks within this document belong to their legitimate owners.
