---
title: "Error calculation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Error calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(
  root.dir = ".."
)
```


## Introduction

In the parameter identification (PI) process, measuring the discrepancy between simulated results and observed data is crucial. `ospsuite.parameteridentification` controls this through two independent settings on `PIConfiguration`: `objectiveType` selects the scoring mode that turns residuals into a single objective value, while `blqMethod` and `blqRemove` control how measurements below the limit of quantification (BLQ) contribute to that score. The error calculation runs within each `PIOutputMapping` and is aggregated across all mappings to drive the optimization.


## Error Model

The error model scores how far the simulation is from the observed data. `objectiveType` selects between two scoring modes, `lsq` (the default) and `mle`.

### LSQ - Least Squares Error

The `lsq` method minimizes the sum of the squared differences between observed and simulated values. It is a widely used, straightforward, and intuitive error calculation for pharmacokinetic modeling.

For a single `PIOutputMapping` with observations $y_i$ measured at times $t_i$ ($i = 1, \dots, n$), the cost is

$$
E = \sum_{i=1}^{n} \left( \hat{y}(t_i) - y_i \right)^2 ,
$$

where $\hat{y}(t_i)$ is the simulated value interpolated to the observation time $t_i$. The total objective driving the optimization is the sum of $E$ over all mappings. Residual weighting and robust adjustment (described below) multiply each residual before it is squared. Under logarithmic scaling the residual is formed between log-transformed values, so it becomes $\log \hat{y}(t_i) - \log y_i$. BLQ substitution (`blqMethod = "lloq"` or `"lloqHalf"`) runs after interpolation and is observed-only: it substitutes $y_i$ when that point's own LLOQ, checked against $y_i$, shows it is BLQ, but never modifies $\hat{y}(t_i)$.

`lsq` is the default and needs no explicit setting, but can be set via `objectiveType`:

```{r, eval=FALSE}
library(ospsuite.parameteridentification)

piConfiguration <- PIConfiguration$new()
piConfiguration$objectiveType <- "lsq"
```

### MLE - Maximum Likelihood Error

The `mle` method minimizes the negative log-likelihood (NLL) of the observations under a Gaussian error model, rather than the raw sum of squares. Each observation is assumed to be normally distributed about the simulated value with a per-observation standard deviation $\sigma_i$:

$$
y_i \sim \text{Normal}\!\left( \hat{y}(t_i), \sigma_i^2 \right),
$$

which gives the negative log-likelihood

$$
\text{NLL} = \frac{n}{2} \ln(2\pi) + \sum_{i=1}^{n} \ln \sigma_i + \sum_{i=1}^{n} \frac{\left( \hat{y}(t_i) - y_i \right)^2}{2 \sigma_i^2} .
$$

Unlike the least-squares total, which is simply $E$ summed over mappings, this $\text{NLL}$ is evaluated once on the observation count $n$ and the weighted sum of squares aggregated across *all* output mappings jointly. It is not a sum of one likelihood value per mapping, because the residual scale is shared across mappings and cannot be concentrated separately within each one; a per-mapping sum of NLLs would not equal the joint likelihood.

The third term is a weighted sum of squares with weights $1/\sigma_i$, so least squares is exactly this negative log-likelihood with the first two terms dropped, which is legitimate only when $\sigma_i$ does not depend on the parameters being estimated. `mle` and `lsq` are therefore two values of one scoring axis rather than two unrelated methods.

`objectiveType` does not take $\sigma_i$ as a separate setting. It is derived from the existing `residualWeightingMethod`:

- **`residualWeightingMethod = "none"`**: the spread is constant across observations and not measured by the data, so $\sigma_i = c / w_i$ for a shape $w_i$ built from the scale factor and the weights already in play, and the unknown scale $c$ is concentrated at its maximum-likelihood value, $c^2 = E / n$ (using the total weighted sum of squares $E$ and the total observation count $n$ aggregated across all mappings). Writing $L_w = -\sum_{i=1}^{n} \ln w_i$ for the summed log of that known shape, substituting the concentrated scale back gives the general form
  $$
  \text{NLL} = \frac{n}{2} \ln(2\pi) + L_w + n \ln(\hat{c}) + \frac{n}{2}, \qquad \hat{c}^2 = \frac{E}{n} .
  $$
  $L_w$ is zero only when every $w_i = 1$ (unit dataset weights, `scaleVar = FALSE`), in which case the formula reduces to the simpler
  $$
  \text{NLL} = \frac{n}{2} \ln(2\pi) + \frac{n}{2} \ln\!\left( \frac{E}{n} \right) + \frac{n}{2} .
  $$
  Forcing `robustMethod = "none"` under `mle` (below) does not make $L_w$ vanish: non-unit dataset weights or `scaleVar = TRUE` still contribute to it, so the general form is the one that applies unless weights are actually unit.
- **`residualWeightingMethod = "error"`**: the observed data supply a measured standard deviation per point (see Residual Weighting below), so $\sigma_i$ is fixed rather than estimated.

Because `lsq` minimizes $E$ directly and, under the constant spread above, $\text{NLL}$ is a strictly increasing function of $E$ for fixed $n$, `mle` and `lsq` reach the **same parameter estimates** under a matching `residualWeightingMethod`. Their objective **values** differ, since one is a sum of squares and the other a likelihood, so objective values are not comparable across `objectiveType`.

```{r, eval=FALSE}
piConfiguration$objectiveType <- "mle"
```

`mle` places three restrictions on the rest of the configuration. `robustMethod` must stay `"none"`, because a robust weight that reaches exactly zero makes the likelihood infinite at every parameter vector rather than merely differently weighted. A dataset weight set via `PIOutputMapping$setDataWeights()` must be strictly positive: under `lsq` a weight of `0` is a legal way to drop a point from the sum of squares, but under a likelihood a zero weight asserts an infinite residual standard deviation for that observation rather than excluding it, so `mle` refuses the configuration rather than silently reinterpreting it; remove the observation instead, or use `blqRemove` if it is below the quantification limit. And with `residualWeightingMethod = "error"`, every observation scored by the fit must carry a usable standard deviation. `lsq` falls back to unit weighting for a row with a missing or invalid error value; `mle` cannot, because that would silently assert a fabricated $\sigma_i = 1$, so it stops with an error naming the offending mapping instead. The same error also reports observations whose value is zero or negative, which is a limitation of the data-error model rather than something missing from the data: the model turns a standard deviation into a weight through the coefficient of variation $\sigma_i / y_i$, which is undefined at a non-positive $y_i$, so a pre-dose measurement of `0` cannot be scored this way however good its reported standard deviation is. Such a row must be removed from the data set, or `residualWeightingMethod` set to `"none"`. Rows the fit never scores at all, those with a missing or infinite value or a negative time, are not reported, because they are discarded before any weight is computed.

Both analytic confidence interval methods, `ciMethod = "hessian"` and `ciMethod = "PL"`, still read the curvature of the objective through least-squares formulas, so neither yet accounts for the likelihood scale under `mle`. Both emit a warning saying so, and their reported widths should not be relied on. `ciMethod = "bootstrap"` is the reliable alternative under `mle`, because it re-optimizes the same objective rather than reading its curvature. One detail of that path is worth knowing: a bootstrap replicate that does not draw a given data set gives every observation of that set a weight of zero, so those observations leave the weighted sum of squares while still counting towards $n$. This shifts the replicate's objective **value**, but not its parameter estimate, since for fixed $n$ the $\text{NLL}$ stays a strictly increasing function of $E$; the interval, which is built from the replicate estimates, is unaffected.


## Handling Data Below the Quantification Limit (BLQ)

A BLQ measurement is known only to lie between zero and the LLOQ. Two `PIConfiguration` settings control how such observations are treated.

`blqMethod` selects how a retained BLQ observation contributes to the error. The substitution methods (`lloq`, `lloqHalf`) are observed-only: a BLQ observation, compared against that point's own LLOQ, is substituted; the simulated prediction is never modified, and a quantifiable observation (at or above the LLOQ) is never touched. This is the standard community/Beal convention; agreement between a censored observation and a prediction below the LLOQ is `m3`'s job, not the substitution methods'.

- **`lloqHalf`** (default): substitute a BLQ observation with half the LLOQ.
- **`lloq`**: substitute a BLQ observation with the LLOQ (PK-Sim default).
- **`none`**: use the stored value unchanged, no substitution.
- **`m3`**: treat the BLQ value as censored and score it by the likelihood that the prediction falls below the LLOQ (Beal's M3 method), keeping all data in the fit for a more statistically robust result. The `blqOptions` field (`linScaleCV`, `logScaleSD`) parameterises this censored contribution. Because censoring is a likelihood operation, `m3` requires `objectiveType = "mle"`.

`blqRemove` selects which BLQ observations enter the fit at all: `none` (default) keeps them all, `always` drops them all, and `trailingSingle` keeps only the first point of each trailing BLQ run.

These settings implement the standard strategies catalogued by Beal (2001), where each method is identified by an M-number:

- **M1** (discard all BLQ): `blqRemove = "always"`.
- **M3** (censored likelihood): `blqMethod = "m3"`.
- **M5** (substitute LLOQ/2): `blqMethod = "lloqHalf"`.
- **M6** (drop trailing BLQ): `blqRemove = "trailingSingle"`.

The publication is listed in the References section.

```{r, eval=FALSE}
piConfiguration$objectiveType <- "mle"
piConfiguration$blqMethod <- "m3"
piConfiguration$blqOptions <- list(linScaleCV = 0.2)
```

Under `m3`, rows flagged as BLQ are excluded from the uncensored objective above, so they are not counted twice, and instead each contributes a censored-likelihood penalty

$$
P_j = -2 \ln \Phi\!\left( \frac{\text{LLOQ}_j - \hat{y}(t_j)}{\sigma_j} \right),
$$

summed over the censored points $j$ and added to the uncensored objective to give the total mapping cost. $\Phi$ is the standard normal CDF and $\ln$ the natural logarithm. The per-point standard deviation $\sigma_j$ is $\lvert \text{linScaleCV} \times \text{LLOQ}_j \rvert$ under linear scaling, or the fixed `logScaleSD` under logarithmic scaling. The default `logScaleSD` is derived from a coefficient of variation of 0.2 as $\sqrt{\ln(1 + \text{CV}^2)} = \sqrt{\ln(1.04)} \approx 0.198$, expressed in natural-log units to match the natural-log transform used by `scaling = "log"`.

The censored penalty $P_j$ is not yet on the same scale as the `mle` objective it is added to: it still draws its $\sigma_j$ from `blqOptions` rather than from the error model above, and it remains a fixed factor of two larger than a plain negative log-likelihood term. `m3` objective values are therefore interim and expected to change again once the two halves are put on a common scale; they should not be treated as a final answer for fitting production data.



## Advanced Error Model Customization

### Residual Weighting 

`PIConfiguration` offers further customization for error calculation through the `residualWeightingMethod`. This option specifies how residuals are weighted. Available options include:

- **`none`** (default): No residual weighting.
- **`error`**: Requires both `yErrorValues` and `yErrorType` to be provided in the observed data `DataSet`. Ideal when individual data points come with variance estimates, allowing for weighting by the inverse of these variances. The `yErrorType` must be a supported OSP Suite error type: `ArithmeticStdDev` or `GeometricStdDev`.
  *Limitation:* When the error field holds population SD for group means, the correct weight is `n / SD²`, not `1 / SD` (where `n` is the group size). Because `DataSet` does not store `n` per time point, this cannot be corrected automatically. Group mean observations will be under-weighted by `1 / √n` relative to the optimal value.

For proportional error handling across outputs of different magnitudes, use `outputMapping$scaling = "log"` instead. Log residuals are dimensionless and automatically comparable across mappings spanning different concentration ranges.

Combining `residualWeightingMethod = "error"` with `outputMapping$scaling = "log"` needs its own conversion, because the residual is then formed between log-transformed values, so the weight must reflect the standard deviation of the log-transformed observation rather than of the observation itself. Treating the observation as lognormal with coefficient of variation $\text{CV}_i = \text{SD}_i / y_i$ gives a log-scale standard deviation of

$$
\sigma_{\log, i} = \sqrt{\ln\!\left(1 + \text{CV}_i^2\right)} ,
$$

which is the conversion used for `ArithmeticStdDev` values. For a `GeometricStdDev` value, the reported quantity is already a multiplicative spread, so the log-scale standard deviation is simply $\ln(\text{GSD}_i)$ directly, with no further conversion. In both cases the weight applied to the log-scale residual is $1 / \sigma_{\log, i}$.

To apply error-based residual weighting, set the `residualWeightingMethod` attribute in `objectiveFunctionOptions` of the `PIConfiguration` object:

```{r, eval=FALSE}
piConfiguration$objectiveFunctionOptions$residualWeightingMethod <- "error"
```


### Robust Residual Calculation 

For enhanced flexibility in handling outliers, ´PIConfiguration´ allows specifying a robust method for residual adjustment. Implementing a robustMethod for residuals directly addresses outlier influence, refining model robustness and ensuring more reliable parameter identification outcomes. Available options include:

- **`none`** (default): No robust adjustments, treating all residuals equally.
- **`huber`**: Minimizes the impact of outliers by combining squared and absolute values, making it less sensitive to extreme deviations.
- **`bisquare`**: Further reduces the influence of outliers by applying a weighting scheme that diminishes the role of residuals as they move away from the median.

To apply robust residual calculation, set the `robustMethod` attribute in `objectiveFunctionOptions` of the `PIConfiguration` object:

```{r, eval=FALSE}
piConfiguration$objectiveFunctionOptions$robustMethod <- "huber" # or "bisquare"
```


### Scaling Impact

The PIOutputMapping$scaling attribute, adjustable to log or lin, crucially shapes the error model's effectiveness and residual calculations. Logarithmic scaling is particularly advantageous for datasets with broad ranges, ensuring proportionate errors and uniform residuals, thereby reducing skewness in error distribution and enhancing model convergence stability.

To change the default ´lin` scaling, set the `scaling` attribute in `PIOutputMapping`:

```{r, eval=FALSE}
outputMapping <- PIOutputMapping$new(quantity = testQuantity)
outputMapping$scaling <- "log"
```


## References

- Beal SL. (2001). Ways to fit a PK model with some data below the quantification limit. *Journal of Pharmacokinetics and Pharmacodynamics*, 28(5), 481-504. [DOI: 10.1023/a:1012299115260](https://doi.org/10.1023/a:1012299115260).

- Fox, J. & Weisberg, S. (2013). Robust Regression. College of Liberal Arts - Statistics, Minneapolis.






