---
title: "Error calculation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Error calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(
  root.dir = ".."
)
```


## Introduction

In the parameter identification (PI) process, measuring the discrepancy between simulated results and observed data is crucial. The `ospsuite.parameteridentification` package offers two primary error models to quantify this difference: the least squares error (`lsq`) and the M3 method (`m3`) for handling data below the quantification limit (BQL).


## Error Models

### LSQ - Least Squares Error

The `lsq` method is a traditional approach that minimizes the sum of the squared differences between observed and simulated values. This method is widely used in pharmacokinetic modeling, offering a straightforward and intuitive error calculation.

With an LLOQ, the `lsq` method compares each observed value with the simulated value at its time as follows:

- A measured value at or above the LLOQ is compared with the simulated value, as without an LLOQ.
- A BQL value, below the LLOQ, has a residual of 0 while the simulated value is below the LLOQ too: both agree that the concentration is below the LLOQ. Above the LLOQ, the residual is the simulated value minus the LLOQ.

The importer stores BQL values as half the LLOQ (LLOQ/2), but this stored value does not enter the residual, so the residual changes continuously with the simulated value, without a jump where it crosses the LLOQ. For example, with an LLOQ of 2, a BQL value has a residual of 0 for a simulated value of 1.5 and of 1 for a simulated value of 3, and a measured value of 3 has a residual of -1.5 for a simulated value of 1.5. With log scaling, the same rule applies to the logarithms of the values and of the LLOQ, so the LLOQ must be positive after the data transformations. Each data set uses its own LLOQ, and the values of a data set without an LLOQ are compared as they are. With y transformations of the observed data, the rule uses the transformed LLOQ (see the Transformations section of `vignette("data-mapping")`).

With `residualWeightingMethod = "error"`, a BQL value that has an error value is weighted by it, calculated with its stored value, as for any other observed value. In `costDetails` of `piResult$toList()`, `ySimulated` is the simulated value and `yObserved` the stored observed value, so for a BQL value, `rawResiduals` is not `ySimulated - yObserved`.

### M3 - Extended Least Squares Error

The `m3` method extends the least squares error by incorporating the likelihood of observed data being below the quantification limit. This method treats BQL data as censored, applying maximum likelihood estimation to include all data in the model fit, thus offering a more accurate and statistically robust approach to handling BQL observations.

Each value below the LLOQ of its own data set is censored. A value at the LLOQ is a measured value and is not censored, and the values of a data set without an LLOQ are not censored either. Every output mapping needs observed data with an LLOQ. The standard deviation of a censored value is `linScaleCV` times its LLOQ with linear scaling, and `logScaleSD` with log scaling. With log scaling, the objective function compares the natural logarithms of the values, so `logScaleSD` is the standard deviation of the natural logarithm. For a coefficient of variation CV, it is `sqrt(log(1 + CV^2))`. The default, `sqrt(log(1 + 0.2^2))` or about 0.198, belongs to a CV of 20%, like the default `linScaleCV` of 0.2. A y offset of the observed data shifts the values and the LLOQ, but not their spread, so with linear scaling the standard deviation is `linScaleCV` times the LLOQ without the y offset, times the absolute y factor. With log scaling, the LLOQ and the stored BQL values, half the LLOQ before the data transformations, must be positive after the transformations, because the BQL values also enter the sum of squared residuals at their stored value. So a y offset must be greater than minus half the LLOQ.

### Configuring Error Models in PI

To apply an error model, set the `targetFunctionType` attribute in `objectiveFunctionOptions` of the `PIConfiguration` object:

```{r, eval=FALSE}
library(ospsuite.parameteridentification)

piConfiguration <- PIConfiguration$new()
piConfiguration$objectiveFunctionOptions$objectiveFunctionType <- "lsq" # or "m3"
```

The error calculation between simulated results and observer datasets within each `PIOutputMapping` directly affects the optimization process, aggregating errors across all mappings.



## Advanced Error Model Customization

### Residual Weighting 

`PIConfiguration` offers further customization for error calculation through the `residualWeightingMethod`. This option specifies how residuals are weighted. Available options include:

- **`none`** (default): No residual weighting.
- **`error`**: Requires both `yErrorValues` and `yErrorType` to be provided in the observed data `DataSet`. Ideal when individual data points come with variance estimates, allowing for weighting by the inverse of these variances. The `yErrorType` must be a supported OSP Suite error type: `ArithmeticStdDev` or `GeometricStdDev`.
  *Limitation:* When the error field holds population SD for group means, the correct weight is `n / SD²`, not `1 / SD` (where `n` is the group size). Because `DataSet` does not store `n` per time point, this cannot be corrected automatically. Group mean observations will be under-weighted by `1 / √n` relative to the optimal value.

For proportional error handling across outputs of different magnitudes, use `outputMapping$scaling = "log"` instead. Log residuals are dimensionless and automatically comparable across mappings spanning different concentration ranges.

To apply error-based residual weighting, set the `residualWeightingMethod` attribute in `objectiveFunctionOptions` of the `PIConfiguration` object:

```{r, eval=FALSE}
piConfiguration$objectiveFunctionOptions$residualWeightingMethod <- "error"
```


### Robust Residual Calculation 

For enhanced flexibility in handling outliers, ´PIConfiguration´ allows specifying a robust method for residual adjustment. Implementing a robustMethod for residuals directly addresses outlier influence, refining model robustness and ensuring more reliable parameter identification outcomes. Available options include:

- **`none`** (default): No robust adjustments, treating all residuals equally.
- **`huber`**: Minimizes the impact of outliers by combining squared and absolute values, making it less sensitive to extreme deviations.
- **`bisquare`**: Further reduces the influence of outliers by applying a weighting scheme that diminishes the role of residuals as they move away from the median.

To apply robust residual calculation, set the `robustMethod` attribute in `objectiveFunctionOptions` of the `PIConfiguration` object:

```{r, eval=FALSE}
piConfiguration$objectiveFunctionOptions$robustMethod <- "huber" # or "bisquare"
```


### Scaling Impact

The PIOutputMapping$scaling attribute, adjustable to log or lin, crucially shapes the error model's effectiveness and residual calculations. Logarithmic scaling is particularly advantageous for datasets with broad ranges, ensuring proportionate errors and uniform residuals, thereby reducing skewness in error distribution and enhancing model convergence stability.

To change the default ´lin` scaling, set the `scaling` attribute in `PIOutputMapping`:

```{r, eval=FALSE}
outputMapping <- PIOutputMapping$new(quantity = testQuantity)
outputMapping$scaling <- "log"
```


## References

- Beal SL. (2001). Ways to fit a PK model with some data below the quantification limit. *Journal of Pharmacokinetics and Pharmacodynamics*, 28(5), 481-504. [DOI: 10.1023/a:1012299115260](https://doi.org/10.1023/a:1012299115260).

- Fox, J. & Weisberg, S. (2013). Robust Regression. College of Liberal Arts - Statistics, Minneapolis.






