---
title: "Error calculation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Error calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_knit$set(
  root.dir = ".."
)
```


## Introduction

In the parameter identification (PI) process, measuring the discrepancy between simulated results and observed data is crucial. `ospsuite.parameteridentification` controls this through two independent settings on `PIConfiguration`: `objectiveFunctionType` selects the error model that scores the fit, while `blqMethod` and `blqRemove` control how measurements below the limit of quantification (BLQ) contribute to that score. The error calculation runs within each `PIOutputMapping` and is aggregated across all mappings to drive the optimization.


## Error Model

The error model scores how far the simulation is from the observed data.

### LSQ - Least Squares Error

The `lsq` method minimizes the sum of the squared differences between observed and simulated values. It is a widely used, straightforward, and intuitive error calculation for pharmacokinetic modeling, and is currently the only available `objectiveFunctionType`.

For a single `PIOutputMapping` with observations $y_i$ measured at times $t_i$ ($i = 1, \dots, n$), the cost is

$$
E = \sum_{i=1}^{n} \left( \hat{y}(t_i) - y_i \right)^2 ,
$$

where $\hat{y}(t_i)$ is the simulated value interpolated to the observation time $t_i$. The total objective driving the optimization is the sum of $E$ over all mappings. Residual weighting and robust adjustment (described below) multiply each residual before it is squared. Under logarithmic scaling the residual is formed between log-transformed values, so it becomes $\log \hat{y}(t_i) - \log y_i$, and BLQ handling can substitute $\hat{y}(t_i)$ or $y_i$ before the difference is taken.

Set it via the `objectiveFunctionType` attribute in `objectiveFunctionOptions`:

```{r, eval=FALSE}
library(ospsuite.parameteridentification)

piConfiguration <- PIConfiguration$new()
piConfiguration$objectiveFunctionOptions$objectiveFunctionType <- "lsq"
```


## Handling Data Below the Quantification Limit (BLQ)

A BLQ measurement is known only to lie between zero and the LLOQ. Two `PIConfiguration` settings control how such observations are treated.

`blqMethod` selects how a retained BLQ observation contributes to the error:

- **`lloqHalf`** (default): substitute the BLQ value with half the LLOQ.
- **`lloq`**: clamp the BLQ value to the LLOQ.
- **`none`**: use the stored BLQ value unchanged.
- **`m3`**: treat the BLQ value as censored and score it by the likelihood that the prediction falls below the LLOQ (Beal's M3 method), keeping all data in the fit for a more statistically robust result. The `blqOptions` field (`linScaleCV`, `logScaleSD`) parameterises this censored contribution.

`blqRemove` selects which BLQ observations enter the fit at all: `none` (default) keeps them all, `always` drops them all, and `trailingSingle` keeps only the first point of each trailing BLQ run.

These settings implement the standard strategies catalogued by Beal (2001), where each method is identified by an M-number:

- **M1** (discard all BLQ): `blqRemove = "always"`.
- **M3** (censored likelihood): `blqMethod = "m3"`.
- **M5** (substitute LLOQ/2): `blqMethod = "lloqHalf"`.
- **M6** (drop trailing BLQ): `blqRemove = "trailingSingle"`.

The publication is listed in the References section.

```{r, eval=FALSE}
piConfiguration$blqMethod <- "m3"
piConfiguration$blqOptions <- list(linScaleCV = 0.2)
```

In the current release the substitution methods (`none`, `lloq`, `lloqHalf`) share the same half-LLOQ behaviour, and `blqRemove` is validated but not yet applied during optimization; these distinctions are introduced in later releases.



## Advanced Error Model Customization

### Residual Weighting 

`PIConfiguration` offers further customization for error calculation through the `residualWeightingMethod`. This option specifies how residuals are weighted. Available options include:

- **`none`** (default): No residual weighting.
- **`error`**: Requires both `yErrorValues` and `yErrorType` to be provided in the observed data `DataSet`. Ideal when individual data points come with variance estimates, allowing for weighting by the inverse of these variances. The `yErrorType` must be a supported OSP Suite error type: `ArithmeticStdDev` or `GeometricStdDev`.
  *Limitation:* When the error field holds population SD for group means, the correct weight is `n / SD²`, not `1 / SD` (where `n` is the group size). Because `DataSet` does not store `n` per time point, this cannot be corrected automatically. Group mean observations will be under-weighted by `1 / √n` relative to the optimal value.

For proportional error handling across outputs of different magnitudes, use `outputMapping$scaling = "log"` instead. Log residuals are dimensionless and automatically comparable across mappings spanning different concentration ranges.

To apply error-based residual weighting, set the `residualWeightingMethod` attribute in `objectiveFunctionOptions` of the `PIConfiguration` object:

```{r, eval=FALSE}
piConfiguration$objectiveFunctionOptions$residualWeightingMethod <- "error"
```


### Robust Residual Calculation 

For enhanced flexibility in handling outliers, ´PIConfiguration´ allows specifying a robust method for residual adjustment. Implementing a robustMethod for residuals directly addresses outlier influence, refining model robustness and ensuring more reliable parameter identification outcomes. Available options include:

- **`none`** (default): No robust adjustments, treating all residuals equally.
- **`huber`**: Minimizes the impact of outliers by combining squared and absolute values, making it less sensitive to extreme deviations.
- **`bisquare`**: Further reduces the influence of outliers by applying a weighting scheme that diminishes the role of residuals as they move away from the median.

To apply robust residual calculation, set the `robustMethod` attribute in `objectiveFunctionOptions` of the `PIConfiguration` object:

```{r, eval=FALSE}
piConfiguration$objectiveFunctionOptions$robustMethod <- "huber" # or "bisquare"
```


### Scaling Impact

The PIOutputMapping$scaling attribute, adjustable to log or lin, crucially shapes the error model's effectiveness and residual calculations. Logarithmic scaling is particularly advantageous for datasets with broad ranges, ensuring proportionate errors and uniform residuals, thereby reducing skewness in error distribution and enhancing model convergence stability.

To change the default ´lin` scaling, set the `scaling` attribute in `PIOutputMapping`:

```{r, eval=FALSE}
outputMapping <- PIOutputMapping$new(quantity = testQuantity)
outputMapping$scaling <- "log"
```


## References

- Beal SL. (2001). Ways to fit a PK model with some data below the quantification limit. *Journal of Pharmacokinetics and Pharmacodynamics*, 28(5), 481-504. [DOI: 10.1023/a:1012299115260](https://doi.org/10.1023/a:1012299115260).

- Fox, J. & Weisberg, S. (2013). Robust Regression. College of Liberal Arts - Statistics, Minneapolis.






