---
title: "Fitting to PK parameter targets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting to PK parameter targets}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```


### Introduction

Parameter identification typically fits a model to a full time profile. In some workflows only a single PK metric is available or relevant as a target: for example, a C_max reconstructed from a human biomonitoring study, or an AUC derived from a non-compartmental analysis. In these cases, optimising the full time-series objective is unnecessary.

`PKOutputMapping` supports this by linking a simulation quantity to a scalar PK metric target. The optimizer adjusts the free parameter until the simulated metric matches the target. Any parameter that has a monotonic effect on the metric can be used.


### Workflow

A PK metric optimization task requires three components:

- A `Simulation` object loaded via `loadSimulation()`.
- A `PIParameters` object wrapping the parameter to optimize.
- One or more `PKOutputMapping` objects, each specifying a simulated quantity, a PK metric, and the target value.

#### Simulation

```{r loadSimulation}
library(ospsuite.parameteridentification)

sim <- loadSimulation(
  system.file("extdata", "Aciclovir.pkml", package = "ospsuite")
)
```

#### Parameter to optimize

The parameter is retrieved by its full path in the simulation. `minValue` and `maxValue` define the search space.

```{r piParameter}
doseParam <- getParameter(
  path = "Events|IV 250mg 10min|Application_1|ProtocolSchemaItem|Dose",
  container = sim
)

piParameterDose <- PIParameters$new(parameters = list(doseParam))
piParameterDose$minValue <- 0.0001 # 100 mg in kg
piParameterDose$maxValue <- 0.001 # 1 g in kg
print(piParameterDose)
```

#### Output mapping

`PKOutputMapping` connects a simulation quantity to a target PK parameter. The `pkParameter` argument must be one of `ospsuite::StandardPKParameter`. The target value must be given in a unit compatible with the PK parameter's dimension.

```{r outputMapping}
plasmaConc <- getQuantity(
  path = "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)",
  container = sim
)

pkOutputMapping <- PKOutputMapping$new(
  quantity = plasmaConc,
  pkParameter = "C_max",
  targetValue = 30,
  targetUnit = "µmol/l"
)
print(pkOutputMapping)
```

If the PK parameter has a different physical dimension than the quantity (e.g. `"AUC_tEnd"` relative to a concentration output), provide `targetUnit` in a unit consistent with what `calculatePKAnalyses()` returns for that PK parameter, which is the base unit of the PK parameter's dimension.


### Running the optimization

Pass the mapping via `pkOutputMappings` to `ParameterIdentification`.

```{r run}
piTask <- ParameterIdentification$new(
  simulations = sim,
  parameters = piParameterDose,
  pkOutputMappings = pkOutputMapping
)

piResult <- piTask$run()
```


### Inspecting results

```{r printResult}
print(piResult)
```

```{r resultDf}
piResult$toDataFrame()[, c(
  "pkParameter",
  "targetValue",
  "targetUnit",
  "achievedValue",
  "estimatedValue",
  "parameterUnit"
)]
```

The data frame contains one row per output mapping with the estimated parameter value (`estimatedValue`), its unit (`parameterUnit`), the achieved PK metric value (`achievedValue`), and the target (`targetValue`, `targetUnit`).

`$toList()` returns the full result including convergence status, number of iterations, elapsed time, and the algorithm used.


### Multiple output mappings

When multiple `PKOutputMapping` objects are supplied, the optimizer minimises the sum of normalised squared differences across all mappings simultaneously.

```{r multiMapping}
pkOutputMappingCmax <- PKOutputMapping$new(
  quantity = plasmaConc,
  pkParameter = "C_max",
  targetValue = 30,
  targetUnit = "µmol/l"
)

pkOutputMappingAuc <- PKOutputMapping$new(
  quantity = plasmaConc,
  pkParameter = "AUC_tEnd",
  targetValue = 2400,
  targetUnit = "µmol*min/l"
)

piTaskMulti <- ParameterIdentification$new(
  simulations = sim,
  parameters = piParameterDose,
  pkOutputMappings = list(pkOutputMappingCmax, pkOutputMappingAuc)
)

piResultMulti <- piTaskMulti$run()

print(piResultMulti)

piResultMulti$toDataFrame()[, c(
  "pkParameter",
  "targetValue",
  "targetUnit",
  "achievedValue",
  "estimatedValue",
  "parameterUnit"
)]
```


By default `ParameterIdentification` uses BOBYQA. A different algorithm or iteration limit can be set via `PIConfiguration`; see `?PIConfiguration` and the [optimization algorithms vignette](optimization-algorithms.html).
