---
title: "Fitting to PK metric targets (QIVIVE)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting to PK metric targets (QIVIVE)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```


### Introduction

Parameter identification typically fits a model to a full concentration-time series. In some workflows only a single PK metric is available or relevant as a target: for example, an in vitro EC50 that must be translated into an equivalent in vivo dose (QIVIVE), or a C_max reconstructed from a human biomonitoring study. In these cases, optimising the full time-series objective is unnecessary.

`PKOutputMapping` supports this by linking a simulation quantity to a scalar PK metric target. The optimizer adjusts the free parameter until the simulated metric matches the target. Any parameter that has a monotonic effect on the metric can be used, although dose is the most common choice for QIVIVE applications.


### Workflow

A PK metric optimization task requires three components:

- A `Simulation` object loaded via `loadSimulation()`.
- A `PIParameters` object wrapping the parameter to optimize.
- One or more `PKOutputMapping` objects, each specifying a simulated quantity, a PK metric, and the target value.

#### Simulation

```{r loadSimulation}
library(ospsuite.parameteridentification)

sim <- loadSimulation(
  system.file("extdata", "Aciclovir.pkml", package = "ospsuite")
)
```

#### Parameter to optimize

The parameter is retrieved by its full path in the simulation. `minValue` and `maxValue` define the search space.

```{r doseParam}
doseParam <- getParameter(
  path = "Events|IV 250mg 10min|Application_1|ProtocolSchemaItem|Dose",
  container = sim
)

piParam <- PIParameters$new(parameters = list(doseParam))
piParam$minValue <- 0.0001  # 100 mg in kg
piParam$maxValue <- 0.001   # 1 g in kg
print(piParam)
```

#### Output mapping

`PKOutputMapping` connects a simulation quantity to a target PK metric. The `pkParameter` argument must be one of `ospsuite::StandardPKParameter`. The target value must be given in a unit compatible with the quantity's dimension.

```{r outputMapping}
plasmaConc <- getQuantity(
  path = "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)",
  container = sim
)

mapping <- PKOutputMapping$new(
  quantity  = plasmaConc,
  pkParameter = "C_max",
  targetValue = 30,
  targetUnit  = "µmol/l"
)
print(mapping)
```

If the PK metric has a different physical dimension than the quantity (e.g. `"AUC_tEnd"` relative to a concentration output), provide `targetUnit` in a unit consistent with what `calculatePKAnalyses()` returns for that metric, which is the base unit of the metric's dimension.


### Running the optimization

Pass the mapping via `pkOutputMappings` to `ParameterIdentification`. Before calling the optimizer, `$run()` executes the simulation once at the parameter's start value and uses the result to estimate a proportionally scaled initial guess, reducing the number of iterations needed.

```{r run}
pi <- ParameterIdentification$new(
  simulations    = sim,
  parameters     = piParam,
  pkOutputMappings = mapping
)

result <- suppressMessages(pi$run())
```


### Inspecting results

```{r printResult}
print(result)
```

```{r resultDf}
result$toDataFrame()
```

The data frame contains one row per output mapping with the estimated parameter value (`estimatedValue`), its unit (`parameterUnit`), the achieved PK metric value (`achievedValue`), and the target (`targetValue`, `targetUnit`).

`$toList()` returns the full result including convergence status, number of iterations, elapsed time, and the algorithm used.


### Target from observed data

When observed concentration-time data is available, `PKOutputMapping` can compute the target metric directly from the dataset instead of requiring a pre-computed scalar. Pass a `DataSet` object to `observedData`; the metric is derived automatically.

```{r observedData, eval = FALSE}
filePath <- system.file(
  "extdata", "Aciclovir_Profiles.xlsx",
  package = "ospsuite.parameteridentification"
)
importConfig <- createImporterConfigurationForFile(filePath)
importConfig$namingPattern <- "{Source}.{Sheet}.{Dose}"
obsData <- loadDataSetsFromExcel(filePath, importConfig, importAllSheets = TRUE)

obs <- obsData$`Aciclovir_Profiles.Vergin 1995.Iv.250 mg`

mappingFromData <- PKOutputMapping$new(
  quantity    = plasmaConc,
  pkParameter = "C_max",
  observedData = obs   # C_max is derived from obs$yValues
)
```

The supported metrics when deriving from a dataset are `C_max`, `C_tEnd`, `t_max`, and `AUC_tEnd`.


### Multiple output mappings

When multiple `PKOutputMapping` objects are supplied, the optimizer minimises the sum of normalised squared differences across all mappings simultaneously.

```{r multiMapping, eval = FALSE}
mappingCmax <- PKOutputMapping$new(
  quantity    = plasmaConc,
  pkParameter = "C_max",
  targetValue = 30,
  targetUnit  = "µmol/l"
)

mappingAuc <- PKOutputMapping$new(
  quantity    = plasmaConc,
  pkParameter = "AUC_tEnd",
  targetValue = 2400,
  targetUnit  = "µmol*min/l"
)

piMulti <- ParameterIdentification$new(
  simulations      = sim,
  parameters       = piParam,
  pkOutputMappings = list(mappingCmax, mappingAuc)
)
```


### Algorithm and configuration

By default, `ParameterIdentification` uses the BOBYQA algorithm, which is efficient for single-parameter problems with smooth objective functions. A different algorithm or iteration limit can be set via `PIConfiguration`.

```{r algorithmConfig, eval = FALSE}
config <- PIConfiguration$new()
config$algorithm <- "HJKB"

piHjkb <- ParameterIdentification$new(
  simulations      = sim,
  parameters       = piParam,
  pkOutputMappings = mapping,
  configuration    = config
)
```

See `?PIConfiguration` and the [optimization algorithms vignette](optimization-algorithms.html) for all available options.


### Non-dose parameters

Any model parameter with a monotonic effect on the target metric can be optimized. The example below uses lipophilicity instead of dose.

```{r nonDoseParam, eval = FALSE}
lipoParam <- getParameter("Aciclovir|Lipophilicity", container = sim)
piLipo <- PIParameters$new(parameters = list(lipoParam))
piLipo$minValue <- -5
piLipo$maxValue <-  5

piLipoTask <- ParameterIdentification$new(
  simulations      = sim,
  parameters       = piLipo,
  pkOutputMappings = mapping
)
```
