---
title: "User guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User guide}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ospsuite.parameteridentification)
knitr::opts_knit$set(
  root.dir = ".." # setting up to package root folder so that the relative paths match between the vignettes and the tests
)
```

### Parameter estimation problems

Biosimulation models include numerical parameters that determine the outputs of the model. Often, the values of these parameters are not known in advance and have to be identified by matching possible model outputs to the observed data. Finding parameter values that best match the observed data is called *parameter identification*.

The `{ospsuite.parameteridentification}` package provides a workflow for setting up such tasks based on existing PKML models, mapping model outputs to observed data, and estimating parameters. Below, three models of increasing complexity are used to demonstrate the functionality of the package.

### PI workflow

The overall workflow is similar to running parameter identification (PI) in PK-Sim or MoBi, see the corresponding [documentation chapter](https://docs.open-systems-pharmacology.org/shared-tools-and-example-workflows/parameter-identification).

To set up a PI task, the user has to define:

-   A set of simulations

    -   You can use multiple simulations in one optimization task. As an example, you can identify a value of the same parameter (e.g., the lipophilicity of the compound) using data for different dosings.

-   Definition of parameters to be identified

-   Mapping of model outputs to observed data

-   Configuration of the optimization task, including selection of algorithm and algorithm options.

#### Simulations

A set of `Simulation` objects. See the documentation of `{ospsuite-r}` on [how to load and adjust simulations](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/load-get.html).

For this example, we will load two instances of the Aciclovir PBPK model provided with the `{ospsuite-r}` package. We want to optimize the lipophilicity and the renal clearance using plasma concentration data gathered after a 10-minute intravenous infusion of aciclovir at doses 250 mg and 500 mg. Lipophilicity will be considered a scenario-independent parameter, i.e., the same value will be applied for both simulations. For the renal clearance, we can assume an inter-individual variability and will optimize the value of this parameter separately for each dose group.

We will start by loading the simulation of the \*.pkml file twice. The dose in this simulation is set to 250 mg. Remember that both `Simulation` objects are created by loading the same \*.pkml file, they represent two independent instances of a simulation.

```{r loadSimulations}
sim_250mg <- loadSimulation(system.file("extdata", "Aciclovir.pkml", package = "ospsuite"))
sim_500mg <- loadSimulation(system.file("extdata", "Aciclovir.pkml", package = "ospsuite"))
```

In the next step, we will retrieve the objects of the application dose parameters and change the dose of the second simulation to 500 mg.

```{r setDose}
# Path to the dose parameter
doseParameterPath <- "Applications|IV 250mg 10min|Application_1|ProtocolSchemaItem|Dose"
# Get the parameter from the first simulation

# Get the instances of the parameters
sim_250mg_doseParam <- getParameter(path = doseParameterPath, container = sim_250mg)
print(sim_250mg_doseParam)
sim_500mg_doseParam <- getParameter(path = doseParameterPath, container = sim_500mg)

# Chage the value to 500 mg
setParameterValues(parameters = sim_500mg_doseParam, values = 500, units = "mg")
print(sim_500mg_doseParam)
```

#### Definition of parameters to be identified

Specification of parameters to be identified is done by creating objects of the class `PIParameters`. A `PIParameter` describes a single model parameter or a group of parameters that will be optimized together, i.e., all model parameters grouped to a `PIParameters` object will have the same value during the optimization. A `PIParameter` also defines the optimization's start and the minimal and maximal values.

If multiple model parameters are linked in one `PIParameters` instance, they can belong to different simulations or come from the same simulation. If you have, for example, two simulations for two individuals with the same compound, you may have one identification parameter, lipophilicity, which is linked to both lipophilicity parameters in the two simulations. At the same time, you can define two identification parameters for the individual reference concentrations of a specific enzyme.

Consider the example above. We will create three `PIParameters` objects - one for the lipophilicity of aciclovir, where the parameters from the two simulations are linked together, and two `PIParameters` for the renal clearance, as they are considered independent.

```{r createPIParameters}
# Creating a PIParameters object with two simulation parameters retrieved from different simulations
piParameterLipo <- PIParameters$new(parameters = list(
  getParameter(path = "Aciclovir|Lipophilicity", container = sim_250mg),
  getParameter(path = "Aciclovir|Lipophilicity", container = sim_500mg)
))

# Creating two separate PIParameters for the renal clearance
piParameterCl_250mg <- PIParameters$new(parameters = getParameter(path = "Neighborhoods|Kidney_pls_Kidney_ur|Aciclovir|Renal Clearances-TS|TSspec", container = sim_250mg))
piParameterCl_500mg <- PIParameters$new(parameters = getParameter(path = "Neighborhoods|Kidney_pls_Kidney_ur|Aciclovir|Renal Clearances-TS|TSspec", container = sim_500mg))
```

Each `PIParameter` has the following properties:

-   `$value`: Current value of the parameter. Corresponds to the value of the `Parameter` object used in the `PIParameters`. If multiple parameters are linked together, the value of the first parameter added is used.

-   `$startValue`: Start value of the parameter(s) used in the identification. By default, the value of the first added parameter. Can be changed.

-   `$minValue`: Minimal allowed value. By default, 0.1-fold of the start value. Can be changed.

-   `$maxValue`: Maximal allowed value. By default, 10-fold of the start value. Can be changed.

-   `$unit`: Unit of the start, min, and max values. WARNING: changing the unit does not update the values! E.g., the default start, min, and max values of the renal clearance parameter are given in `1/min`:

```{r}
print(piParameterCl_250mg)
```

Setting the unit to `1/h` will cause the identification to start at `0.94 1/h`, while the current value is still `0.94 1/h` or `0.94 1/h * 60 min = 56.47 1/h`:

```{r}
piParameterCl_250mg$unit <- ospUnits$`Inversed time`$`1/h`
print(piParameterCl_250mg)
```

We will set the boundaries for lipophilicity to [-10, 10], and for renal clearance to [0, 10] `1/min`:

```{r}
piParameterLipo$minValue <- -10
piParameterLipo$maxValue <- 10

piParameterCl_250mg$minValue <- 0
piParameterCl_250mg$maxValue <- 10
piParameterCl_250mg$unit <- ospUnits$`Inversed time`$`1/min`

piParameterCl_500mg$minValue <- 0
piParameterCl_500mg$maxValue <- 10
piParameterCl_500mg$unit <- ospUnits$`Inversed time`$`1/min`

print(piParameterLipo)
print(piParameterCl_250mg)
print(piParameterCl_500mg)
```

#### Mapping of model output to observed data

In order to calculate the error of the model, simulation outputs must be linked to observed data. One simulation output can be linked to multiple observed data sets in this framework. The error will be calculated for each pair of simulated output and linked data set and added up for the total error. The mapping is done by creating an object of the class `PIOutputMapping`. Each object links a single [`Quantity`](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/Quantity.html) of a simulation to multiple [`DataSet`](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/DataSet.html) objects.

In the first step, we will load two observed data sets from the Excel file provided with the package. Read the article [Observed data](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/observed-data.html) from the `{ospsuite}` documentation for details.

```{r loadDataSet}
filePath <- system.file("extdata", "Aciclovir_Profiles.xlsx", package = "ospsuite.parameteridentification")

# Create importer configuration for the file
importConfig <- createImporterConfigurationForFile(filePath = filePath)
# Set naming patter
importConfig$namingPattern <- "{Source}.{Sheet}.{Dose}"
# Import data sets
obsData <- loadDataSetsFromExcel(xlsFilePath = filePath, importerConfigurationOrPath = importConfig, importAllSheets = TRUE)

print(names(obsData))
```

The data sets hold concentration data of aciclovir in venous blood; the corresponding simulation output has the path:

```{r}
simOutputPath <- "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"
```

We will create two objects of the `PIOutputMapping` class, one for each dose group:

```{r}
outputMapping_250mg <- PIOutputMapping$new(quantity = getQuantity(path = simOutputPath, container = sim_250mg))

outputMapping_500mg <- PIOutputMapping$new(quantity = getQuantity(path = simOutputPath, container = sim_500mg))
```

and link simulation results to the corresponding observed data:

```{r}
outputMapping_250mg$addObservedDataSets(obsData$`Aciclovir_Profiles.Vergin 1995.Iv.250 mg`)
outputMapping_500mg$addObservedDataSets(obsData$`Aciclovir_Profiles.Vergin 1995.Iv.500 mg`)

print(outputMapping_250mg)
```

The scaling of the output mapping defines how the residuals between simulated and observed data are calculated. For details, see the article error-functions. For concentration data, we will change the scaling to logarithmic:

```{r}
outputMapping_250mg$scaling <- "log"
outputMapping_500mg$scaling <- "log"
```

All data sets (simulated results and observed data) within a `PIOutputMapping` can be transformed by applying offsets and scaling factors following the same logic implemented in the [`DataCombined` class](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/data-combined.html#transformations).

#### Configuration

The configuration of a PI includes the selection of an algorithm, setting algorithm settings, and some further options. A PI configuration is defined in a `PIConfiguration` object and can be re-used by multiple PI tasks. If no user-defined configuration is provided to a PI task, a default one is created with the following properties:

```{r}
piConfiguration <- PIConfiguration$new()
print(piConfiguration)
```

-   `printEvaluationFeedback` (logical) indicates if the objective function value should be printed at each iteration

-   `simulationRunOptions`: Object of type [`SimulationRunOptions`](https://www.open-systems-pharmacology.org/OSPSuite-R/reference/SimulationRunOptions.html) that will be passed to simulation runs. If `NULL`, default options are used.

-   `targetFunctionType` (possible values are listed in `ospsuite.parameteridentification::ObjectiveFunctionOptions`) indicates which objective function should be used to calculate the difference between observed data and simulated curve. See the article `vignette("error-calculation")` for more information.

-   `algorithm` (possible values: `r paste0(ospsuite.parameteridentification:::Algorithms, collapse = ", ")`) indicates which of the optimization algorithms should be used. Algorithm options can be specified as a list and passed to `piConfiguration$algorithmOptions`. Supported algorithms and their options are described in `vignette("optimization-algorithms")`.

The default settings are listed in `AlgotitmOptions_XYZ`, where `XYZ` is the algorithm's name. For example, the default settings for the `BOBYQA` algorithm are stored in `AlgotitmOptions_BOBYQA`:

```{r}
print(AlgotitmOptions_BOBYQA)
```

E.g. to set the maximal number of function evaluations to 1500 for the `BOBYQA` algorithm, we can use the following code:

```{r}
defOptions <- AlgotitmOptions_BOBYQA
defOptions$maxFunctionEvaluations <- 1500
piConfiguration$algorithmOptions <- defOptions
```

#### Running a parameter identification task

After the simulation, identification parameters, mappings of observed and simulated data, and the configuration are defined, they are used in an instance of the `ParameterIdentification` class:

```{r include=TRUE, eval=TRUE}
piTask <- ParameterIdentification$new(
  simulations = list(sim_250mg, sim_500mg),
  parameters = list(piParameterLipo, piParameterCl_250mg, piParameterCl_500mg),
  outputMappings = list(outputMapping_250mg, outputMapping_500mg),
  configuration = piConfiguration
)
```

We first can create time profiles comparing simulation results with observed data with the current parameter values. A separate time profile is created for each `PIOutputMapping`.

```{r}
piTask$plotResults()
```

NOTE: the figures are created with the current values of the parameters, not the start values!

We can now run the PI and print the results:

```{r}
piResults <- piTask$run()
print(piResults)
```

The output of the optimization run is:

-   `taskResults$par`: a vector of point estimates for each of the parameters
-   `taskResults$lwr` and `taskResults$upr`: vectors with lower bounds and upper bounds for the 95% confidence interval for each of the parameters
-   `taskResults$cv`: a vector with the coefficient of variation (standard deviation over point estimate, in percents) for each of the parameters
-   `taskResults$value` is the objective function value at the point estimate
-   `taskResults$elapsed` (seconds) is the wall time it took to run the main optimization routine, excluding the hessian calculations
-   `taskResults$nrOfFnEvaluations` is the number of times the objective function was evaluated in the main optimization routine, excluding the hessian calculations

The optimal values are directly applied to the `Parameter` objects used in the PI.

#### Diagnostics

To assess the goodness of the fit, plot the time profiles after the optimization:

```{r}
piTask$plotResults()
```

The quality of the fit can be assessed by the following criteria:

-   The individual time profile is close to the observed data points.
-   The predicted-vs-observed plot is close to the diagonal.
-   The residuals are randomly distributed above and below zero.

The residuals only above or only below zero indicate an overprediction or an underprediction. Another set of parameters will likely result in a better fit.
